Supplementary Materials for \"COATi: statistical pairwise alignment of protein coding sequences\

### 2024-04-18 alignpair_letter.tex
Draft of revision 2

\usepackage[letterpaper,margin=1in]{geometry}

@@

To address these issues, we present COATi, a statistical, codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data. COATi allows users to reduce the amount of discarded data while generating more accurate sequence alignments. COATi can infer indels both within and between codons, leading to improved sequences alignments. We applied COATi to a dataset containing orthologous protein-coding sequences from humans and gorillas, and conclude that 41\% of indels occurred between codons, agreeing with an earlier study in rats and mice. We also applied to semi-empirical benchmark alignments and find that COATi outperforms several popular alignment programs via several measures.

@@

To address current limitations of alignment software to accurately align protein-coding sequences, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates evolutionary models for protein-coding sequences and is robust to artifacts present in modern genomic data sets.

@@

We have developed a statistical alignment program using finite-state transducers called COATi. On empirical data, COATi produced alignments that provided reasonable biological inferences. On semi-empirical, simulated benchmarks COATi was more accurate than the other aligners/alignment strategies that we tested: ClustalΩ's amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model. Unlike standard codon or amino-acid alignment strategies, COATi supports aligning protein-coding sequences using all three phases of gaps. Consistent with \cite{taylor2004occurrence} and \cite{zhu2022profiling}, our COATi results estimate that only 41\% indels in protein-coding sequences between humans and gorillas are phase-3 indels. This indicates that alignment strategies that only support phase-3 gaps in protein-coding sequences are sub-optimally aligning the 59\% of indels that are phase-1 or phase-2.

In addition to supporting all three phases of gaps, COATi's other advantage over competing aligners in our tests is at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligner's default settings. Different data sets may show different results. However, because COATi uses a statistical model with biologically meaningful parameters, users can easily optimize COATi by using parameters estimated by previous studies or estimating parameters themselves \citep[e.g.][]{zhu2022profiling}. This contrasts with other aligners that have limited ability to customize parameters and/or use parameters without biological interpretations.

@@

We can look at distances between alignments generated by different aligners to understand the diversity of approaches use by our alignment protocols. Aside from the cluster of different COATi methods, we detected no other clusters in our analysis (Supplemental Materials Fig.~S7). Different aligners tended to spread out from one another and in different directions from the benchmark. This indicates that each protocol that we utilized approached alignment inference from a different section of ``aligner space''. While each aligner, except ClustalΩ, performed well on its own gap patterns, MACSE also did well on ClustalΩ and MAFFT gap patterns, and PRANK also did well on ClustalΩ patterns. COATi did well on all patterns.

COATi uses finite-state transducers to align pairs of sequences using a codon-aware statistical model, and the present study has shown that COATi offers a biologically significant improvement over other methods when aligning pairs of sequences separated by short evolutionary distances and in the presence of genomic artifacts. While COATi has multiple models, we have focused on evaluating the utility of COATi's complex finite-state-transducer for estimating pairwise alignments. In other work, we have begun to explore application of COATi to more divergent sequences and the accuracy of the marginal approximation of the COATi FST \citep{garcia2023dissertation}.

@@

>>>>

\usepackage[letterpaper,margin=0.75in]{geometry}

@@

To address these issues, we present COATi, a statistical, codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data. COATi allows users to reduce the amount of discarded data while generating more accurate sequence alignments. COATi can infer indels both within and between codons, leading to improved sequences alignments. We applied COATi to a dataset containing orthologous protein-coding sequences from humans and gorillas, and conclude that 41\% of indels occurred between codons, agreeing with previous work in rodents. We also applied to semi-empirical benchmark alignments and find that COATi outperforms several popular alignment programs on several measures of alignment quality and accuracy.

@@

To address current limitations of alignment software to accurately align protein-coding sequences, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates evolutionary models for protein-coding sequences and is robust to artifacts present in modern genomic data sets. We find that COATi generates biologically reasonable alignments and downstream inferences when applied to empirical data. We also find that COATi generates more accurate alignments when applied to semi-empirical benchmarks.

@@

We have developed a statistical alignment program using finite-state transducers called COATi. On empirical data, COATi produced alignments that provided reasonable biological inferences. On semi-empirical benchmarks COATi was more accurate than the other aligners/alignment strategies that we tested: ClustalΩ's amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model. Unlike standard codon or amino-acid alignment strategies, COATi supports aligning protein-coding sequences using all three phases of gaps. Consistent with \cite{taylor2004occurrence} and \cite{zhu2022profiling}, our COATi results indicate that only 41\% indels in protein-coding sequences between humans and gorillas are phase-3 indels. Therefore alignment strategies that only support phase-3 gaps in protein-coding sequences are sub-optimally aligning the 59\% of indels that are phase-1 and phase-2.

In addition to supporting all three phases of gaps, COATi's advantage over the other aligners that we tested is at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligners' default settings. Different data sets may show different results. However, because COATi uses a statistical model with biologically meaningful parameters, users can easily optimize COATi by using parameters estimated by previous studies or estimating parameters themselves \citep[e.g.][]{zhu2022profiling}. This contrasts with other aligners that have limited ability to customize parameters and/or use parameters without biological interpretations.

@@

We can look at distances between alignments generated by different aligners to understand the diversity of approaches use by our alignment protocols. Aside from the cluster of different COATi methods, we detected no other clusters in our analysis (Supplemental Materials Fig.~S7). Different aligners tended to spread out from one another and in different directions from the benchmark. This indicates that each protocol that we utilized approached alignment inference from a different section of ``aligner space''. While each aligner, except ClustalΩ, performed well on its own gap patterns, MACSE also did well on ClustalΩ and MAFFT gap patterns, and PRANK did well on ClustalΩ patterns. COATi did well on all patterns.

@@

The present study has shown that COATi offers a biologically significant improvement over other methods. COATi produces more accurate alignments of protein-coding sequences and better downstream inferences of sequence divergence, selection, and indel processes. While here we have focused on the challenge of aligning pairs of sequences from sister species that may contain genomic artifacts, in other work, we have begun to explore application of COATi to more divergent sequences and evaluate in detail the accuracy of the marginal approximation of the COATi FST \citep{garcia2023dissertation}.
